---
output: html_notebook
---
# Setup Notebook

```{r setup}
knitr::opts_knit$set(root.dir = '..')
devtools::install_github("tidyverse/ggplot2") #version with stat_qq_line
```

```{r}
# Load requirements
library(dplyr)
library(tidyr)
library(ggplot2)
library(readr)
```

# About this noteboook
This r markdown notebook contains the figures and tables used in reports for this project.

# Milestone Report
## Background
The total spending on primary and secondary education is comparable to the federal defense budget.

* [US Defense Budget data](https://www.statista.com/statistics/272473/us-military-spending-from-2000-to-2012/)
* [Local Education Agency Finance Survey Data Fiscal 2014](https://nces.ed.gov/ccd/data/zip/sdf14_1a_flat.zip)

```{r}
# Read in fiscal 2014 financial data
data <- read.table(unz('data/finance_fiscal2014.txt.zip', 'finance_fiscal2014.txt'), sep='\t', header=TRUE, quote='', comment.char='')

# Create dataframe to represent defense spending
defense_budget <- data.frame(Type=c('Defense', 'Defense', 'Defense'), Source=c('Federal', 'State', 'Local'), Spend=c(609910000000, 0, 0))

# Wrangle education and defense data into a single dataframe
data <- data %>%
  select(TFEDREV, TSTREV, TLOCREV) %>%
  summarise(Federal=sum(as.numeric(TFEDREV)), State=sum(as.numeric(TSTREV)), Local=sum(as.numeric(TLOCREV))) %>%
  gather('Source', 'Spend') %>%
  bind_cols(Type=c('Education', 'Education', 'Education')) %>%
  bind_rows(defense_budget) %>%
  mutate(Spend = Spend / 1000000000)

# Create stacked bar plot
ed_v_def <- ggplot(data, aes(x=Type, y=Spend, fill=Source)) + 
  geom_bar(stat="sum", position = "stack") +
  labs(x='Type', y='Spend ($ Billion)')

# Print stacked bar plot
ed_v_def

# Save stacked bar plot
ggsave('output/plots/ed_v_def.png', ed_v_def)

```

6 Percent of Local Education Agencies in 2009-2010 closed within 5 years
* 381,248 Students
* $4,032,423,000 Spending

```{r, fig.width=6, fig.height=1.5}
# Read in 2009-2010 data
data <- read.csv('output/wrangled_data_ii.csv')

# Calculate percentage of school districst from 2009-2010 that were closed in 2014-2015
closed_districts <- nrow(filter(data, exist_five_years == 'False'))/nrow(data)
print(paste('School Districts from 2009-2010 that were closed in 2014-2015:', closed_districts, sep=' '))

# Create stacked bar plot, print bar plot, save bar plot
closed_districts <- data %>% select(exist_five_years) %>% group_by(exist_five_years) %>% summarize(count=n())
closed_districts_plot <- ggplot(closed_districts, aes(1, count, fill=exist_five_years, label=count)) +
  geom_bar(stat="identity") +
  coord_flip() +
  geom_text(size = 3, position = position_stack(vjust = 0.5)) +
  scale_fill_discrete(name="Operational \n2014-2015") +
  theme(axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        text = element_text(size=15)) +
  ylab('Number of 2009-2010 School Districts')
closed_districts_plot
ggsave('output/plots/closed_districts.png', closed_districts_plot)

affected_students <- data %>%
  filter(exist_five_years == 'False') %>%
  select(total_students, total_revenue)

print(paste('Total students in closed school distrcts:', sum(affected_students$total_students, na.rm=TRUE), sep=' '))
print(paste('Total spending on closed school districts:', sum(affected_students$total_revenue, na.rm=TRUE), sep=' '))
```


Even without labels, clustering identified school districts more likely to close
* Recall (test set): 0.398
* Precision (test set): 0.411

Recall and Precision were evaluated on a test set during analysis. I did a sanity check by evaluating recall and precision on the entire 2009-2010 dataset.

```{r}
# Read in cluster results on 2009-2010 data
data <- read.csv('output/cluster_results.csv')

# Summarize data by cluster
aggregated_data <- data %>%
  group_by(cluster) %>%
  summarize(total_school_districts = n(), 
            closed_school_districts=sum(label=='True')) %>%
  select(cluster, total_school_districts, closed_school_districts) %>%
  mutate(closed_school_districts = closed_school_districts/total_school_districts) %>%
  arrange(desc(closed_school_districts))

# Create bar plot
clusters_plot <- ggplot(aggregated_data, aes(x=reorder(cluster, -closed_school_districts),
                                             y=closed_school_districts,
                                             fill=total_school_districts,
                                             label=total_school_districts)) +
  geom_bar(stat="identity") + 
  coord_cartesian(ylim = c(0, 0.6)) +
  geom_text(size = 3, position = position_dodge(width=1), vjust=-0.25) +
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        text = element_text(size=15)) +
  ylab('Proportion Closed 2014-2015') +
  xlab('Clusters') + 
  scale_fill_continuous(name="Total School Districts \n2009-2010")
clusters_plot
ggsave('output/plots/clusters_plot.png', clusters_plot)

# Sanity Check Recall and Precision
recall <- nrow(data %>% select(label, cluster) %>% filter(label=='True', cluster %in% c(3, 1))) / 
  nrow(data %>% select(label) %>% filter(label=='True'))
precision <- nrow(data %>% select(label, cluster) %>% filter(label=='True', cluster %in% c(3, 1))) / 
  nrow(data %>% select(label, cluster) %>% filter(cluster %in% c(3, 1)))
```

